State of the Practice for Lattice Boltzmann Method Software

### 2022-02-15 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updates to SOP for LBM, start of digging deeper paper

@@

\begin{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
	the pain points for developers working on LBM software projects?
	(Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
	do the pain points of developers from LBM compare to the pain points
	for research software in general? (Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For LBM
	developers what specific practices are taken to address the pain point
	concerns? (Section~\ref{Sec_AddressConcerns})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
	What, if any, existing research software best practice, not currently used
	by LBM software, could address the concerns identified in
	\rqref{RQ_Concerns}). (Section~\ref{Sec_AddressConcerns})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Future}:] What practices
	could be employed by future LBM developers to address the pain points found
	in~\rqref{RQ_PainPoints} and the concerns identified
	in~\rqref{RQ_Concerns}? (Section~\ref{Sec_AddressConcerns})

@@

	\item[P\refstepcounter{pnum}\thepnum \label{P_Correctness}:]
	\textbf{Ensuring Correctness} Difficulties with ensuring correctness were
	noted by several developers. The TCLB developer commented that the amount of
	testing data needed was sometimes problematic, since free testing services
	do not offer adequate facilities for large amounts of data, which means
	in-house testing solutions are needed.  Other research software domains
	point to different problems with testing, as follows: i)
	\citep{PintoEtAl2018} mention the problem of insufficient testing; ii) the
	survey of \citep{HannayEtAl2009} shows that more developers think testing is
	important than the number that believe they have a sufficient understanding
	of testing concepts; and, iii) \citep{HannayEtAl2009, KanewalaAndBieman2013,
	KellyEtAl2011, WieseEtAl2019} point to the oracle problem, which
	occurs in research software when we do not have a means to judge the
	correctness of the calculated solutions.  Since LBM developers did not
	allude to the oracle problem in our conversation, we believe that they have
	developed techniques to work around the oracle problem.

@@

	\item[P\refstepcounter{pnum}\thepnum \label{P_Documentation}:]
	\textbf{Quality of Documentation} One of the consequences of pain point
	\ppref{P_LackDevTime} is sometimes inadequate documentation. The
	documentation for LBM is challenging because of the complexity of the
	underlying CFD theory and the differing backgrounds of users. Poor research
	software documentation is also mentioned by others \citep{PintoEtAl2018,
	WieseEtAl2019} and the problem also arises with non-research software
	\citep{LethbridgeEtAl2003}.  Recommendations on the development of research
	software state that developers should critically evaluate their own
	development processes in terms of quality assurance and comply with
	international standards for software documentation
	\citep{KaterbowAndFeulner2018}.

@@

\section{LBM Developer Identified Concerns} \label{Sec_Concerns}

\subsection{Correctness and Verifiability}

Interviews with developers confirmed that these software packages are developed
by domain experts with backgrounds in physics, mathematics and mechanical
engineering. \wss{Could point to surveys that have shown the same thing.}  It
was noted in these interviews that some of the developers do not have formal
software engineering education. \wss{Can point to other studies to support
this.} Some of the development teams include computer scientists. Despite a lack
of visible domain documentation and a resulting lower surface correctness and
verifiability score, it is clear that some of the software packages were
developed by teams with significant domain knowledge.

Interviews suggested a more frequent use of both unit testing and continuous
integration in the development processes than what was observed from the initial
survey. For example, OpenLB, pyLBM, and TCLB use such methods during
development, despite this not being explicitly clear from an analysis of the
material available online. \wss{Add this to threats to validity.}  The
correctness and verifiability of such packages is not measured well using
surface analysis.

Several developers alluded to difficulty with testing the correctness of large
numbers of features, and some even manually tested program output, as opposed to
using automated testing. The use of well defined unit testing tools could
decrease the time spent testing some feature.  \wss{link this to
recommendations}

\subsection{Usability}

Interviews with developers revealed several usability issues. Some users have
misunderstood the boundaries of LBM and CFD, and have combined or applied
methods that are not physically sound. Sometimes users have applied LBM to
poorly defined or inappropriate fluid dynamics problems. For example, they may
wish to model flow through or around a structure despite having limited
information about the structure or its environment, and having little previous
knowledge of CFD. The users do not realize the limitations of the methods, of
the software, and do not understand the requirements to properly model a problem
with the software. As the developer of TCLB noted, such software packages are
not designed to be used ``out of the box'' in a plug and play fashion, and it
could take months or more to set up the CFD problems correctly. Developers of
some software packages, including ESPResSo, mitigated this by editing the source
code to prevent users from ``combining methods that are not physically sound
together'', and by updating the documentation to better inform users of LBM
limitations, and of the requirements to properly model appropriate problems,
including what algorithms and parameters to use.  \wss{Could point to this for
recommendations related to documentation.  Could point to this for a generative
approach that won't allow inadmissible combinations, or at least includes an
explicit warning.}

Usability concerns were also raised for installation, understanding how to
maneuver the interfaces and understanding how to set up and run models. These
issues are addressed by various user support models, including frequently asked
questions sections on the software package websites, user guides, and hardware
and software requirements specifications.  \wss{Could point to there being room
for improvement here, especially for requirements specification.}

\subsection{Maintainability}

Although not always visible from the publicly available artifacts, interviews
with developers revealed that most projects have a defined process for accepting
contributions from team members. The packages rarely get contributions from
outside developers, but the process would be similar as for the aforementioned
group.  In general, contributions are made through GitHub, and are then reviewed
and pulled by lead developers, often with consultation with a group of core
developers depending on the organizational model. Continuous integration is part
of the process for some packages. \wss{Recommendation to increase.  Successful
projects use it, a reasonable next step for projects that have not yet
incorporated CI.}  \wss{There is definitely a great deal of repetition.}

Some developers noted that their software package does not have well defined
contributing guide in the repository, but it might be a good idea to add one in
the near future. They would be happy to see contribution from outside of their
organization, but currently this does not happen.  \wss{A comment was made at
the recent DOE workshop that you get new users through documentation and
tutorials.  Maybe highlight the example of Musubi?  Should trace back to this
for the recommendations part.  \citep{Fogel2005} gives advice on having a
successful project, also the web article (if needed).}

Furthermore, maintainability has been addressed by increasing source code
modularity, reducing duplicate information, and improving abstraction by
developing well defined interfaces. This was noted by the developers of ESPResSo
and pyLBM. Several software packages have had sections of their code base
redeveloped with languages that the developers felt are more understandable and
readable, and that are better supported, such as Python. Data structures have
also been redeveloped and storage has been improved. A developer of pyLBM
mentioned that the geometries and models of their system had been ``decoupled'',
using abstraction and modularization of the source code, to make it ``very easy
to add [new] features''.  \wss{Highlights the advantage of design.  Other
projects can look to these examples.  Maybe cite the Anshu paper about Flash?
Dune?  and point to the advantages of design.}

The developer of TCLB mentioned that their package had two sets of code, for
executing the models on the CPU and GPU, and that maintenance was decreased by
introducing macros, a practice which became a common part of the development
process.

\subsection{Modifiability}

Software packages were not quantitatively measured for modifiability. In this
project we asked developers to comment on modifiability when we interviewed
them. Specifically, we asked if ease of future changes to the system, modules,
and code blocks was considered when designing the software. We also asked if any
measures had been taken to ensure the ease of future changes. All of the
developers that were interviewed noted that the ease of future changes was
considered and that measures to ensure it had been taken, including requiring
the separation of software components in the source code architecture.  \wss{A
good approach for other developers to adopt - part of recommendations.  Trace
back to this comment.}  A high degree of code modularity and abstraction was
noted by developers as a measure to ensure the ease of future changes. This can
be ensured by separating components and hiding information behind well defined
interfaces. The developer of ESPResSo also noted that some of the code base was
transitioned from C to C++, which could ease modifiability of that software
package. The developer of TCLB noted that their software package was designed to
allow for the addition of some LBM features, but changes to major aspects of the
system would be difficult. For example, ``implementing a new model will be an
easy contribution'', but changes to the ``Cartesian mesh...will be a
nightmare''. Furthermore, the package was designed with flexible data structures
and storage in mind.  \wss{example suggests identifying likely changes.}

Developers were asked to comment on the obstacles in their development process.
The developer of ESPResSo noted that a lot of their source code had been written
with a specific application in mind, and that there is too much coupling between
components. Addressing this issue, possibly by making the design review process
more transparent (Section~\ref{secVisib}), would help with code modifiability
and reusability.  \wss{Traceback for recommendations.}

\subsection{Understandability}

Software developers noted that they believe users have generally found their
packages to be understandable. \wss{Confusion between understandability and
usability - threat to validity example.}  The interviewed developer of ESPResSo
commented that some users have attempted to run physically incompatible LBM
methods, and the solution was to edit the code to prevent such combinations, as
well as to update the documentation to prevent misunderstanding the methods.
Similarly, a developer of pyLBM noted that some users had issues setting up
parameters for LBM schemes. The solution to this was to update the interface
where these parameters are set, as well as to add functionality to test the
stability of the parameters. A developer of OpenLB noted that some users lack
the background knowledge to easily model fluid dynamics problems using their
software. A frequently asked questions section was added to their package
website to help users find answers to common questions. The package also has
detailed documentation, including guides and usage requirements specification,
to better help users understand the software. \wss{Peter - the discussion here
is more about user usability than the understandability of the code for
developers. Did any of the developers discuss developer understandability?  The
confusion between understandability and usability was also a confusion during
the interviews Ao conducted.} \pmm{The developer of ESPResSo did comment on
developer understandability, specifically for developers that are/were part of
their peer review process. He commented that code architecture is documented,
and that code is made to be readable, and that these help peer reviews of the
code.}

\wss{I moved this from the visibility section, since it is more about
understandability.}  Developers were also asked how documentation fits into their
development process. Several developers noted that developer documentation plays
an important role in familiarizing potential contributors to the software system
architecture. Without the guidance that the documentation provides it would be
unlikely that contributions would pass the peer review process.

\subsection{Traceability}

Developers noted the importance of documentation for both the users and
developers of their software. New features are always added to the
documentation. The developers use documentation to stay up to date on the status
of the software package, and to help expand features, like computational models
or algorithms. This is necessary so that the coding standard for these models is
kept consistent for new developers. \wss{Peter - this also sounds like
discussion that is relevant for understandability.} \pmm{The points here sound
like they could work for both. The definition of traceability notes that `This
quality refers to keeping track of information as it changes forms or relates
between artifacts'. So the point `New features are always added to the
documentation' could be reworded to fit that part of the definition of
traceability. As noted above, this is similar to the point about some packages
having including detailed change logs. Same with `The developers use
documentation to stay up to date on the status of the software package, and to
help expand features'. I agree that at least some of this paragraph could be in
understandability, especially the first and second sentences.}

The importance of documentation for both users and developers was stressed
throughout the interviews. However, it was noted several times that a lack of
time and funding has a negative affect on the documentation. Most of the
developers are scientific researchers evaluated on the scientific papers that
they produce. Writing and updating documentation is something that is done in
their free time, if that time arises. Sometimes it is a last priority for the
developers. Finding ways to hasten updating documentation would increase the
frequency of such updates and benefit both users and developers.  \wss{Same
thing has come up in papers.  I'll try to find a reference for this.  This is
also something we can point to for recommendations related to code generation.}
\wss{Peter - this also sounds like understandability related discussion?  I
wonder if we should drop Traceability as a quality that we assessed?  There
doesn't seem to be much content specific to traceability.}  \pmm{Based on the
comments I just added above, I think we can keep at least a small part of what
is here in traceability, rewording parts to highlight the importance of change
logs. Some of the content in this section should indeed be moved to
understandability.} The developer of OpenLB noted the use of documentation
generators like Doxygen. It would be advisable for more projects to use such
automatic document generation tools, since some projects do not do this.
\wss{We can refer back to this for the recommendations.}

For the developers interviewed, visibility is not currently a priority.
Visibility is not particularly necessary because the teams are fairly small.
Although they accept outside contributors, these are rare and the teams remain
tight-knit.  Most of the teams interviewed have all members at the same
institution, although one team is spread out internationally. \wss{Peter, can
you say which team is international, so that we can add this detail to the
paper?} \pmm{The interviewed developer of OpenLB noted that there were tasks for
local and international developers when discussing their software development
model. He also noted that they have frequent local meetings and annual
international meetings.}  For teams of only a few developers, new features or
major changes are discussed with the entire team. Projects with more than a few
developers have lead developer roles. These lead developers review potential
additions to the software.

Although visibility is not a priority, positive steps are being undertaken. For
instance, all interviewed teams used GitHub, with its visibility enhancing
features of branching, issue tracking, insights on contributors, the blame
feature, wikis, pull requests, etc. Continuous integration, as used by ESPResSo,
makes projects even more visible, since contributors can explicitly see the bar
that must be crossed for code to be incorporated into the master branch.
ESPResSo also uses peer review, which improves visibility by explicitly asking
for feedback from other developers.

Previously mentioned concerns with maintainability, modifiability, and
understandability  could be improved by further improving visibility.  For
future sustainability, projects will likely have to expand their development
teams by recruiting more external developers. One component of recruiting
developers is transparency in the development process, including the process for
making a contribution.  \wss{in recommendations could point out what teams could
do to get more external contributors, and potentially a longer life for their
project.}

@@

\subsection{Unambiguity}

Software packages were not quantitatively measured for unambiguity. We asked
developers if they thought that the current documentation can clearly convey all
necessary knowledge to the users, and if they had taken any steps to ensure
clarity. The developer of ESPResSo noted that their documentation was meant for
users that are already familiar with the underlying physics and CFD methods.
These concepts are not explained in detail within the documentation. Users
should acquire this knowledge from suitable external sources. The documentation
focuses on how to technically use the software package, and includes a user
guide and tutorial walk through of how to set up and run a simulation. With this
in mind, the developer believes that their documentation is in reasonable shape
for users with a minimum knowledge of the underlying physics. If new users have
technical questions these can be addressed in further revisions of the
documentation. New developers should find that the documentation is reasonably
clear and useful. Information that is missing, like detailed explanations of
dependencies, is referenced in the documents.

The developer of pyLBM also noted that their documentation was in reasonable
shape, but that they ``need more [user] feedback to improve [it]''. They also
noted that they believe a lack of knowledge of the underlying physics and CFD
concepts can be an issue for some users. This information can be referenced in
the documentation, but it is not something that the documentation needs to
detail.

Answers \rqref{RQ_Recommend}.  Recommendations for Addressing Developer
Concerns.  This section answers the research questions listed in
Section~\ref{secIntro} using the quantitative data presented in
Section~\ref{AHPresults}, qualitative data from developer interviews
(Section~\ref{SecSurvey}), additional data from software package repositories
and artifacts, and domain expert feedback.

@@

\subsection{Common Practices} \label{currentPractices}

The practices that arose frequently.  Compare and contrast to what is done with
research software in general.  The following points regarding software quality
should be considered when developing LBM software packages. These points are
based on developer interviews, SCS literature, what was found to have worked for
packages that were designated as high quality in this assessment.  Our
recommendations are not lists of what should have been done in the past, or what
should be done now; they are just suggestions for consideration in the future.
\wss{Rework this section to highlight what the top projects are doing.  Point to
checklists for other advice?}  Highlights of common practices.  Put in context.

Conclusion that LBM software common practice is similar to the common practice
elsewhere.  Maybe point out where LBM is ahead.  Use results from common
artifacts, common process etc. to support conclusion.  (Continuous integration,
version control, README, Req etc. - point to Table 8.)

\begin{itemize}
	\item Surface Reliability. Surface Usability. The packages should include
	detailed tutorials, including dependencies, expected output, and any
	additional supplementary documentation that may be required. This was done
	by many packages, including waLBerla, Palabos, MechSys, LUMA, and pyLBM.
	\wss{Comment from ASCR attendee that success with open source requires good
	tutorials and documentation - is this published anywhere?  I think I can
	find this recommendation in other papers.}
	\item Surface Usability. Unambiguity. Surface Understandability. Provide a
	detailed user manual. It should identify elements of user interfaces, and
	identify all requirements to model a system. Thirteen packages have a user
	manual. ESPResSo has a well detailed manual. \wss{Include this with
	tutorial?  Find papers that recommend this kind of documentation?}
	\item Surface Usability. Include user hardware and software requirements
	documentation. Hardware requirements are rarely listed in the software
	packages that were assessed. Only those offering GPU processing (ESPResSo,
	lbmpy, lettuce, pyLBM, Sailfish, TCLB) mentioned any hardware requirements.
	On the other hand, some sort of software requirements were available for all
	packages, even if only some dependencies or a compatible operating system
	were mentioned.  \wss{This is a key part of reproducibility.}
    \item Not much of a process - find papers that support this - papers
    recommend "Define clear and transparent contribution, governance and
    communication processes" JimenezEtAl2017
\end{itemize}

\subsection{Best Practices} \label{bestPractices}

practices by the best measured.  Point to existing checklist and development
guides.  This section should only mention things that are new.

\begin{itemize}
	\item Surface Correctness and Verifiability. Development teams should
	include both domain experts and experienced software developers. This
	suggestion is based on developer comments.
	\item Surface Correctness and Verifiability. Surface Understandability.
	Maintainability.  Modifiability. Reusability.  Modularize the source code,
	separate components, hide information behind well defined interfaces. This
	is suggested in SCS literature, and in developer comments.  \wss{one of
	Wilson's best practices.}
	\item Surface Usability. Surface Understandability. Unambiguity. State
	appropriate fluid dynamics problems that the software is designed to model
	in the documentation, and explicitly state the limits of the software. This
	is done in the ESPResSo user guide.
	\item Surface Usability. Provide documentation that details the background
	theory information, or provide a reference to such information. This was
	done by several top ranked packages, including ESPResSo, Ludwig, LUMA, and
	OpenLB.
	\item Surface Usability. Identify expected user characteristics. LIMBES,
	Ludwig, laboetie, and Palabos did this. The importance of specifying user
	characteristics is discussed in \citep{smith2007requirements}.
	\item Surface Usability. Keep all documentation in one location. This was
	done by top ranked packages.
	\item Maintainability. Have a defined process for accepting contributions,
	and make public documentation for making contributions to the project. The
	top five ranked software packages include information on how to contribute. 
	\item Modifiability. Consider flexibility of data structures and data
	storage in the design stage. The package pyLBM redeveloped data structures
	to ease future changes.
	\item Reusability. Provide API documentation, if applicable. Only one
	(ESPResSo) of the top five ranked packages provided API documentation.
	\item Surface Understandability. Adopt a coding standard and document it in
	artifacts, including some examples. A coding standard needs to be part of
	continuous integration. Only LUMA, HemeLB, and Musubi explicitly identify
	coding standards. 
	\item Traceability. Provide a developer's guide to help orient developers to
	the artifacts, source code, and system architecture so that they can better
	document changes. Top ranked package OpenLB has a developer guide. \wss{How
	is this related to traceability?}
	\item Visibility and Transparency. Summarize the development process that is
	used. Provide information on how new users can contribute. Identify the
	development model by name (waterfall, agile, etc.), if appropriate.  Seven
	of the software packages have some artifacts outlining the general
	development process. Eleven packages have information on how to contribute. 
	\item Visibility and Transparency. Consider peer review processes to assess
	contributions and ensure the tracking of information. High ranked package
	ESPResSo uses a peer review process for contributions. 
	\item Reproducibility. Test output against automatically calculated or known
	correct results. High ranked package ESPResSo does this.
	\item Reproducibility. Automate the testing of output. Run tests after all
	code changes. High ranked package ESPResSo does this.
	\item Reproducibility. Consider peer review processes and task based
	inspection to assess contributions. High ranked package ESPResSo does this.
	\item Unambiguity. Consider asking users for feedback on the documentation.
	The developer of pyLBM noted that such feedback would be appreciated.

\end{itemize}

practices mentioned in literature, but not practiced (if any).  Start with the
practices that aren't derived from developer interviews.

\begin{itemize}
	\item Surface Correctness and Verifiability. Use a requirements
	specification document. This is suggested in SCS literature, and several of
	the top ranked packages had such a document or reference to theory manuals.
	A potential template is presented in \citep{smith2005new}. 
	\item Surface Usability. Do not duplicate artifact information. This is
	suggested as good software development practice. Duplicate information is
	difficult to maintain. 
	\item Maintainability. Visibility and Transparency. Reproducibility. Keep
	artifacts updated. This is suggested as good software maintenance practice.
	\item Maintainability. Eliminate code duplication. This is suggested as good
	software development practice. Duplicate code is difficult to maintain.
	\wss{No support for this one or citation, either find or remove it.}
	\item Modifiability. Consider future source code modifiability as early as
	the design stage of development. This is suggested as good software
	development practice. \wss{citation, support, or remove.}
	\item Reusability. Document module interfaces in design and developer
	documentation. This is suggested as good software development practice.
	\wss{citation or support}
	\item Traceability. Update all relevant documentation when a change to the
	software is made. This is suggested as good software development practice.
	\wss{support, cite or remove}
	\item Unambiguity. Documentation must state all technical requirements and
	software dependencies. A missing dependency was a frequent cause of fault
	conditions during this assessment.
	\item Unambiguity. Provide a table of symbols in the developer's guide that
	maps to names used in the source code. This is suggested as good software
	development practice.

\end{itemize}

\subsection{Future Practices} \label{futurePractices}

Answers \rqref{RQ_Future}.  practices for the future.

@@

@@

\subsection{Highlighted Recommendations} \label{highlightedrecommendations}

@@

The following recommendations improve software quality along multiple
attributes, and provide the greatest return on investment:

\begin{itemize}
	\item Provide a detailed user manual. It should identify elements of user
	interfaces, and identify all requirements to model a system. Thirteen
	packages have a user manual. ESPResSo has a well detailed manual.
	\item State appropriate fluid dynamics problems that the software is
	designed to model in the documentation, and explicitly state the limits of
	the software. This is done in the ESPResSo user guide.
	\item Include detailed tutorials, including dependencies, expected output,
	and any additional supplementary documentation that may be required. This
	was done by many packages, including waLBerla, Palabos, MechSys, LUMA, and
	pyLBM.
	\item Keep all documentation in one location. This was done by top ranked
	packages.
	\item If possible, consider using popular user-friendly software languages
	like Python. Especially consider this for parts of the source code that is
	likely to be modified or reviewed by users. ESPResSo and Sailfish use Python
	to address several software qualities, including maintainability,
	modifiability and usability.
	\item Modularize the source code, separate components, hide information
	behind well defined interfaces. This is suggested in SCS literature, and in
	developer comments.
	\item Include descriptive error messages where appropriate. This was done by
	most packages that encountered a fault.
	\item Summarize the development process that is used. Provide information on
	how new users can contribute. Seven of the software packages have some
	artifacts outlining the general development process. Eleven packages have
	information on how to contribute.
	\item Consider peer review processes and task based inspection to assess
	contributions. High ranked package ESPResSo does this.
	\item Use continuous integration tools (Bamboo, Jenkins, and Travis CI) and
	processes during development. This was done by top ranked packages ESPResSo
	and Ludwig.
	\item Use project management tools, including change and version control
	tools (GitHub, GitLab, CVS), collaboration tools (GitHub, GitLab), and
	document generation tools (Doxygen, Drasil, Sphinx). This was done by all of
	the top five ranked packages.
\end{itemize}

Ensuring software quality can take significant time. An inability to develop
high quality documentation due to time constraints could be mitigated in the
future by the use of automatic document generation tools like Drasil. The Drasil
Framework consists of a collection of Domain Specific Languages (DSL) for
capturing scientific documents, structures, and computing knowledge, and then
transforming this knowledge into relevant software artifacts without having to
manually duplicate knowledge into multiple artifacts \citep{zhao2018}.
\wss{Yuzhi is not the best reference to use for the Drasil work.}

@@

- add table of features - include Mohammed columns
- add new program that wasn't measured - means measuring 24, not 23
- update simple stats for other projects
- compare to existing survey results (do this in Ao's too)
- read more of the existing survey studies, as highlighted by Carver paper
- data on Mendeley

- add a comparison to other studies
- add conclusions related to how LBM compares to other fields, and the "average" of all SC

@@

- value of tools to enforce good behavior of developers, like R, like MOOSE infrastructure

- change RQs - make them more about understanding the range of development
practices, and identifying projects that are successful.  Also identify room for
improvement, as identified by the developers themselves, and as identified by
us.  Compare the state of the practice of LBM to the state of practice of other
domains.  Where does LBM stand relative to other domains.

- read \citet{WieseEtAl2019} (Naming the Pain in Developing Scientific Software)

@@

- add links or citations to all of the packages (could add them to the summary table)

@@

- discuss Wilson papers ``Best Practices'' \citep{WilsonEtAl2014} and ``Good
Enough Practices'' \citep{WilsonEtAl2016}

@@

- JimenezEtAl2017 talks about Define clear and transparent contribution,
governance and communication processes

>>>>

\newcounter{bpnum} %best practice number
\newcommand{\bpthepnum}{BP`'\thebpnum}
\newcommand{\bpref}[1]{BP\ref{#1}}

@@

\begin{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
	the pain points for developers working on LBM software projects?
	(Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
	do the pain points of developers from LBM compare to the pain points
	for research software in general? (Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For LBM
	developers what specific best practices are taken to address the pain points
	and software quality concerns? (Section~\ref{Sec_AddressConcerns})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
	What research software development practice could potentially address the
	pain point concerns identified in \rqref{RQ_PainPoints}).
	(Section~\ref{Sec_Recommendations})
\end{enumerate}

@@

	\item[P\refstepcounter{pnum}\thepnum \label{P_Correctness}:]
	\textbf{Ensuring Correctness} Difficulties with ensuring correctness were
	noted by several developers. Several developers alluded to difficulty with
	testing the correctness of large numbers of features, and some manually
	tested program output, as opposed to using automated testing. The TCLB
	developer commented that the amount of testing data needed was sometimes
	problematic, since free testing services do not offer adequate facilities
	for large amounts of data, which means in-house testing solutions are
	needed.  Other research software domains point to the following problems
	with testing: i) \citep{PintoEtAl2018} mention the problem of insufficient
	testing; ii) the survey of \citep{HannayEtAl2009} shows that more developers
	think testing is important than the number that believe they have a
	sufficient understanding of testing concepts; and, iii)
	\citep{HannayEtAl2009, KanewalaAndBieman2013, KellyEtAl2011, WieseEtAl2019}
	point to the oracle problem, which occurs in research software when we do
	not have a means to judge the correctness of the calculated solutions.  The
	LBM experience seems to overlap with i and ii, but the LBM developers did
	not allude to problem iii (the oracle problem) in our conversation; we
	believe that they have developed techniques to work around the oracle
	problem.

@@

	\item[P\refstepcounter{pnum}\thepnum \label{P_Documentation}:]
	\textbf{Quality of Documentation}  The importance of documentation for both
	users and developers was stressed throughout the interviews. However, it was
	noted several times that a lack of time and funding (\ppref{P_LackDevTime})
	has a negative affect on the documentation. Most of the developers are
	scientific researchers evaluated on the scientific papers that they produce.
	Writing and updating documentation is something that is done in their free
	time, if that time arises. Inadequate research software documentation is
	also mentioned by others \citep{PintoEtAl2018, WieseEtAl2019} and the
	problem also arises with non-research software \citep{LethbridgeEtAl2003}.
	Recommendations on the development of research software state that
	developers should critically evaluate their own development processes in
	terms of quality assurance and comply with international standards for
	software documentation \citep{KaterbowAndFeulner2018}.

@@

This section summarizes the best practices from LBM developers that are taken to
address the pain points mentioned in Section~\ref{painpoints}.  The main source
of information is the qualitative data from developer interviews
(Section~\ref{SecSurvey}).  The practices summarized in this section can
potentially be emulated by the LBM software packages that do not currently
follow them.  Moreover, these practices may also provide examples that can be
followed by other research software domains.

\subsection{Design For Change} \label{Sec_DesForChange}

Increasing source code modularity, reducing duplicate information,
and improving abstraction by developing well defined interfaces. This was noted
by the developers of ESPResSo and pyLBM.  A developer of pyLBM mentioned that
the geometries and models of their system had been ``decoupled'', using
abstraction and modularization of the source code, to make it ``very easy to add
[new] features''.  \wss{Highlights the advantage of design.  Other projects can
look to these examples.  Maybe cite the Anshu paper about Flash? Dune? and point
to the advantages of design.} Addresses maintainability.  The developer of TCLB
mentioned that their package had two sets of code, for executing the models on
the CPU and GPU, and that maintenance was decreased by introducing macros, a
practice which became a common part of the development process.

A high degree of code modularity and abstraction was noted by developers as a
measure to ensure the ease of future changes. This can be ensured by separating
components and hiding information behind well defined interfaces. The developer
of ESPResSo also noted that some of the code base was transitioned from C to
C++, which could ease modifiability of that software package. The developer of
TCLB noted that their software package was designed to allow for the addition of
some LBM features, but changes to major aspects of the system would be
difficult. For example, ``implementing a new model will be an easy
contribution'', but changes to the ``Cartesian mesh...will be a nightmare''.
\wss{example suggests identifying likely changes.}

Modularize the source code, separate components, hide information behind well
defined interfaces. This is suggested in SCS literature, and in developer
comments.  \wss{one of Wilson's best practices.}

\begin{itemize}
	\item Modifiability. Consider flexibility of data structures and data
	storage in the design stage. The package pyLBM redeveloped data structures
	to ease future changes.
	\item Reusability. Provide API documentation, if applicable. Only one
	(ESPResSo) of the top five ranked packages provided API documentation.
\end{itemize}

\subsection{Verification and Validation} \label{Sec_VnV}

Surface Correctness and Verifiability. Development teams should
include both domain experts and experienced software developers. This
suggestion is based on developer comments.

Visibility and Transparency. Consider peer review processes to assess
contributions and ensure the tracking of information. High ranked package
ESPResSo uses a peer review process for contributions.

@@

\begin{itemize}
\item Reproducibility. Test output against automatically calculated or known
correct results. High ranked package ESPResSo does this.
\item Reproducibility. Automate the testing of output. Run tests after all code
changes. High ranked package ESPResSo does this.
\item Reproducibility. Consider peer review processes and task based inspection
to assess contributions. High ranked package ESPResSo does this.
\end{itemize}

\subsection{Documentation} \label{Sec_RecommendDoc}

Finding ways to hasten updating documentation would increase the
frequency of such updates and benefit both users and developers.  \wss{Same
thing has come up in papers.  I'll try to find a reference for this.  This is
also something we can point to for recommendations related to code generation.}
\wss{Peter - this also sounds like understandability related discussion?  I
wonder if we should drop Traceability as a quality that we assessed?  There
doesn't seem to be much content specific to traceability.}  \pmm{Based on the
comments I just added above, I think we can keep at least a small part of what
is here in traceability, rewording parts to highlight the importance of change
logs. Some of the content in this section should indeed be moved to
understandability.} The developer of OpenLB noted the use of documentation
generators like Doxygen. It would be advisable for more projects to use such
automatic document generation tools, since some projects do not do this.
\wss{We can refer back to this for the recommendations.}

\begin{itemize}
\item Usability. Understandability. Unambiguity. State appropriate fluid
dynamics problems that the software is designed to model in the documentation,
and explicitly state the limits of the software. This is done in the ESPResSo
user guide.
\item Usability. Provide documentation that details the background theory
information, or provide a reference to such information. This was done by
several top ranked packages, including ESPResSo, Ludwig, LUMA, and OpenLB.
\item Usability. Identify expected user characteristics. LIMBES, Ludwig,
laboetie, and Palabos did this. The importance of specifying user
characteristics is discussed in \citep{smith2007requirements}.
\item Usability. Keep all documentation in one location. This was done by top
ranked packages.
\item Usability. Do not duplicate artifact information. This is suggested as
good software development practice. Duplicate information is difficult to
maintain. 
\end{itemize}

\wss{Will likely remove this section, but should find a way to say the good
things that ESPResSo does.} Although visibility is not a priority, positive
steps are being undertaken. For instance, all interviewed teams used GitHub,
with its visibility enhancing features of branching, issue tracking, insights on
contributors, the blame feature, wikis, pull requests, etc. Continuous
integration, as used by ESPResSo, makes projects even more visible, since
contributors can explicitly see the bar that must be crossed for code to be
incorporated into the master branch. ESPResSo also uses peer review, which
improves visibility by explicitly asking for feedback from other developers.

\wss{in recommendations could point out what teams could do to get more external
contributors, and potentially a longer life for their project.}

\section{Recommendations for Future Practices} \label{Sec_Recommendations}

Answers \rqref{RQ_Recommend}.  practices for the future.  Probably the easiest
approach is to list the practice and then the PPs that it helps.  Our
recommendations are not lists of what should have been done in the past, or what
should be done now; they are just suggestions for consideration in the future.

@@

Surface Correctness and Verifiability. Use a requirements
specification document. This is suggested in SCS literature, and several of
the top ranked packages had such a document or reference to theory manuals.
A potential template is presented in \citep{smith2005new}.

\subsection{Grow the Number of Contributors}

Maintainability. Have a defined process for accepting contributions, and make
public documentation for making contributions to the project. The top five
ranked software packages include information on how to contribute.

Papers that talk about how to do open source.

@@

Interviews suggested a more frequent use of both unit testing and continuous
integration in the development processes than what was observed from the initial
survey. For example, OpenLB, pyLBM, and TCLB use such methods during
development, despite this not being explicitly clear from an analysis of the
material available online.  The correctness and verifiability of such packages
is not measured well using surface analysis.

@@

@@

Ensuring software quality can take significant time. An inability to develop
high quality documentation due to time constraints could be mitigated in the
future by the use of automatic document generation tools like Drasil. The Drasil
Framework consists of a collection of Domain Specific Languages (DSL) for
capturing scientific documents, structures, and computing knowledge, and then
transforming this knowledge into relevant software artifacts without having to
manually duplicate knowledge into multiple artifacts \citep{SzymczakEtAl2016}.

@@

- data on Mendeley

@@

- value of tools to enforce good behaviour of developers, like R, like MOOSE infrastructure

@@

@@

@@
